---
title: "Peer Assessement 1"
author: "André Reis"
date: "06/11/2015"
output: html_document
---

## Loading and preprocessing the data  

```{r echo=TRUE}
activity_monitoring <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?  

```{r echo=TRUE}
require(ggplot2)
steps_per_day <- sapply(split(activity_monitoring,activity_monitoring$date),function(x){sum(x[,1],na.rm = TRUE)})
mean_steps <- mean(steps_per_day)
median_steps <- median(steps_per_day)
qplot(steps_per_day,geom="histogram",binwidth = 1000,main="Number of steps taken per day",xlab="steps per day")
```

The average number of steps taken per day is `r sprintf("%.2f", mean_steps)` and the median is `r median_steps`.

## What is the average daily activity pattern?  

```{r echo=TRUE}
mean_interval <- sapply(split(activity_monitoring,as.factor(activity_monitoring$interval)),function(x){mean(x[,1],na.rm = TRUE)})
intervals <- as.integer(names(mean_interval))
qplot(intervals,mean_interval,geom="line",main="Average daily activity pattern",xlab="5-minute interval",ylab="average number of steps")
max_interval <- names(sort(mean_interval,decreasing=TRUE)[1])
```

The 5-minute interval, that on average across all the days in the dataset, contains the maximum number of steps is `r max_interval`.

## Imputing missing values

```{r echo=TRUE}
number_NAs <- dim(activity_monitoring[!complete.cases(activity_monitoring),])[1]
```

The total number of missing values in the dataset is `r number_NAs`.

The mean number of steps for each 5-minute interval was used to impute the missing values.

```{r echo=TRUE}
for (i in 1:nrow(activity_monitoring)) {
   if (is.na(activity_monitoring[i,1])) {
       activity_monitoring[i,1] <- mean_interval[as.character(activity_monitoring[i,3])]
   }
}
steps_per_day <- sapply(split(activity_monitoring,activity_monitoring$date),function(x){sum(x[,1],na.rm = TRUE)})
mean_steps <- mean(steps_per_day)
median_steps <- median(steps_per_day)
qplot(steps_per_day,geom="histogram",binwidth = 1000,main="Number of steps taken per day",xlab="steps per day")
```

The average number of steps taken per day is `r sprintf("%.2f", mean_steps)` and the median is `r sprintf("%.2f", median_steps)`. These values differ from the estimates from the first part of the assignment, but not by much. In general, imputing missing data did not have a big impact on the estimates of the total daily number of steps.

## Are there differences in activity patterns between weekdays and weekends?

```{r echo=TRUE}
for (i in 1:nrow(activity_monitoring)) {
    if (weekdays(as.Date(activity_monitoring[i,2])) %in% c("Saturday","Sunday")) {
        activity_monitoring[i,4] <- "weekend"
    } else {
        activity_monitoring[i,4] <- "weekday"
    }
}
activity_monitoring[,4] <- as.factor(activity_monitoring[,4])
mean_interval_week <- as.data.frame(t(sapply(split(activity_monitoring,list(as.factor(activity_monitoring$interval),activity_monitoring[,4])),function(x){c(mean(x[,1],na.rm = TRUE),x[1,3],as.factor(x[1,4]))})))
colnames(mean_interval_week) <- c("number_steps","interval","day_week")
mean_interval_week$day_week <- as.factor(mean_interval_week$day_week)
levels(mean_interval_week$day_week) <- c("Weekdays","Weekends")
ggplot(mean_interval_week, aes(x=interval, y=number_steps)) + geom_line() + facet_grid(~day_week) + xlab("5-minute interval") + ylab("Number of steps")
```

The highest number of steps occurs during the the weekdays, but in the weekends, in general, the mean number of steps throughout most of the intervals tend to be higher.